Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-11-14 SEIR.tex
add in comments

The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
This term can be arrived at by considering a single infected individual.
Assume this individual makes contact with any other individual in the population at a constant rate, $\kappa$, in units of contacts per day.
Further, assume the population is \emph{well-mixed}, meaning the individual is equally likely to come into contact with any other individual.
For a randomly selected individual, the probability that they are susceptible is $s(t)$.
Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Therefore, in expectation, the infected individual makes contact with susceptible individuals at a rate of $\kappa s(t)$.
Assume that there is a constant probability of transmission upon contact between an infected and susceptible individual, $p_{si}$.
The expected rate of infections caused by this infected individual is therefore $\kappa s p_{si}$.
The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa p_{si}$.
Therefore, the instantaneous number of infections per infected individual is $\beta s(t)$.
The overall proportion of the population instantaneously infected is $\beta s(t)i(t)$.

@@

@@

$K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
$\beta_{ij}$ is then estimated, although often with a parsimonious model.
A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
Informally, it is how vulnerable to infection individuals in stratum $i$ are.
Formally, it is the probability that a susceptible individual in stratum $i$ will be infected upon contact with an infectious individual, relative to a individual in the reference stratum.
Similarly, $\beta^\text{inf}_i$ is known as the \emph{relative infectiousness} of stratum $i$.
Informally, it is how likely individuals in stratum $i$ are to pass on the infection.
Formally, it is the relative risk of an infectious individual in stratum $i$ infecting a susceptible individual upon contact, relative to an individual in the reference stratum.
These measures could be biological, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygenic than adults.

@@

Based on prior work\todo{cite}, I assume that all individuals have the same \emph{infectiousness} but that there is a difference in \emph{susceptibility}.
Therefore, $\beta^\text{inf}_{j} = 1$ for all $j$.
The prior literature suggests that children are less susceptible than adults, but otherwise there is little difference in susceptibility.
Therefore, I set $\beta^\text{susc}_i$ to 1 for adult-aged strata and $\beta^\text{susc}_i = \beta_c$ for child-aged strata.

\subsection{Geographic structure} \label{SEIR:sec:geography}

@@

Between-age-group contacts rates have previously been inferred from a combination of contact surveys, time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
I use the outputs from this analysis as $\matr{M}$.
To summarise, baseline rates and contexts of contacts between age groups are taken from the contact survey POLYMOD~\autocite{mossongSocial}.
How these contacts relate to time use is taken from the UK Time Use Survey\todo{cite time use survey}.
How the population spent their time each week during the pandemic is taken from the Google mobility data\todo{cite mobility data}, which measured the locations of Android users over the period of interest.
\todo[inline]{is the above paragraph too hand wavy?}

@@

\section{Observation model} \label{SEIR:sec:observation}

@@

A $l$-phase EC distribution is a convolution of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
A schematic of an EC distribution and its parameters is shown in \cref{SEIR:fig:EC}.

@@

For any short period of time when $\vec{s}$ is approximately constant, the system of ODEs can be solved analytically to produce exponential growth.
In fact, all the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same initial exponential growth rate, $\psi$.
If $\vec{s} = 1$ then the basic reproduction number can be linked to the growth rate of this epidemic as follows~\autocites{birrellRealtimea}{wearingAppropriate}:
\begin{align}
    \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
\end{align}

@@

\begin{landscape}
\begin{table}
\begin{tabular}{l c l l}
    Parameter description & Symbol & Distribution & Comment \\
    \hline \\
    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & From a paper \\
    Mean infectious period & $1/\gamma$ & 4 days (fixed) & From a paper \\
    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity} & \\
    Initial proportion infected & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
    Relative susceptibility of children & $\beta_c$ & $\LN(-0.4325, 0.1174)$ &  \\
    Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
    Standard deviation of random walk & $\sigma$ & $\Exponential(80)$ & Weakly informative penalised model complexity prior~\autocite{simpsonPenalising} \\
    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Weakly informative
\end{tabular}
\todo[inline]{Find citation for children, add priors for $\pi$}
\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
\label{SEIR:table:priors}
\end{table}
\end{landscape}

>>>>

\todo[inline]{double check correct terms for the below, especially rates vs instantaneous proportion and per unit time vs per day}
The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
This term can be arrived at by considering a single infected individual.
Assume this individual makes contact with any other individual in the population at a constant rate, $\kappa$, in units of contacts per day.
Further, assume the population is \emph{well-mixed}, meaning the individual is equally likely to come into contact with any other individual.
For a randomly selected individual, the probability that they are susceptible is $s(t)$.
Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Therefore, in expectation, the infected individual makes contact with susceptible individuals at a rate of $\kappa s(t)$.
Assume that there is a constant probability of transmission upon contact between an infected and susceptible individual, $p_{si}$.
The expected rate of infections caused by this infected individual is therefore $\kappa s p_{si}$.
The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa p_{si}$.
Therefore, the instantaneous number of infections per infected individual is $\beta s(t)$.
The overall proportion of the population instantaneously infected is $\beta s(t)i(t)$.

@@

To relax the assumption of a well-mixed population, the population can be stratified by some characteristic.

\todo[inline]{start with a generic intro on structured populations and use of matrices, then give age as a motivating example, finally specifics go elsehwere}

@@

$K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
% $\beta_{ij}$ is then estimated, although often with a parsimonious model\todo{ref / give examples / delete on how age-specfic susceptibility / transmissibility happens}.
A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
Informally, it is how vulnerable to infection individuals in stratum $i$ are.
Formally, define an $a_0$ and $a'_0$ and enforce $\beta^\text{susc}_{a_0}=1$ for identifiability.
Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
\todo{clarify notation and explaantion here of reference class}
it is the probability that a susceptible individual in stratum $i$ will be infected upon contact with an infectious individual, relative to a individual in the reference stratum.
Similarly, $\beta^\text{inf}_i$ is known as the \emph{relative infectiousness} of stratum $i$.
Informally, it is how likely individuals in stratum $i$ are to pass on the infection.
Formally, it is the relative risk of an infectious individual in stratum $i$ infecting a susceptible individual upon contact, relative to an individual in the reference stratum.
These measures could be biological, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygenic than adults.

@@

\todo[inline]{move the below into a context-specific section}
Based on prior work\todo{cite}, I assume that all individuals have the same \emph{infectiousness} but that there is a difference in \emph{susceptibility}.
Therefore, $\beta^\text{inf}_{j} = 1$ for all $j$.
The prior literature suggests that children are less susceptible than adults, but otherwise there is little difference in susceptibility.
Therefore, I set $\beta^\text{susc}_i$ to 1 for adult-aged strata and $\beta^\text{susc}_i = \beta_c$ for child-aged strata.

\subsection{Geographic structure} \label{SEIR:sec:geography}
\todo[inline]{remove this section and just explain what I'm doing in this case in an implementation section}

To allow for regional hetereogeneity....
This assumes non-interaction between regions...

@@

Between-age-group contacts rates have previously been inferred from a combination of contact surveys, time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
I use the outputs from this analysis as $\matr{K}$ in equation\todo{ref where this contact matrix fits in}.
To summarise, baseline rates and contexts of contacts between age groups are taken from the contact survey POLYMOD~\autocite{mossongSocial}.
How these contacts relate to time use is taken from the UK Time Use Survey\todo{cite time use survey}.
How the population spent their time each week during the pandemic is taken from the Google mobility data\todo{cite mobility data}, which measured the locations of Android users over the period of interest.
\todo[inline]{is the above paragraph too hand wavy?}

@@

\section{Observation model} \label{SEIR:sec:observation}
\todo[inline]{Be clear this section is to fit to CIS data}

@@

\todo[inline]{need more explanation here: what is a phase-type distribution, be less terse in the following text}
A $l$-phase EC distribution is a convolution of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
A schematic of an EC distribution and its parameters is shown in \cref{SEIR:fig:EC}.

@@

For any short period of time when $\vec{s}$ is approximately constant, the system of ODEs can be solved analytically to produce exponential growth.
In fact, all the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same initial exponential growth rate, $\psi$.
\todo{move R definition for each model into where they are introduced}
If $\vec{s} = 1$ then the basic reproduction number can be linked to the growth rate of this epidemic as follows~\autocites{birrellRealtimea}{wearingAppropriate}:
\begin{align}
    \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
\end{align}

@@

\begin{landscape}
\begin{table}
\begin{tabular}{l c l l}
    Parameter description & Symbol & Distribution & Comment \\
    \hline \\
    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & From a paper \\
    Mean infectious period & $1/\gamma$ & 4 days (fixed) & From a paper \\
    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
    Initial proportion infected & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
    Relative susceptibility of children & $\beta_c$ & $\LN(-0.4325, 0.1174)$ &  \\
    Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
    Standard deviation of random walk & $\sigma$ & $\Exponential(80)$ & Weakly informative penalised model complexity prior~\autocite{simpsonPenalising} \\
    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Weakly informative
\end{tabular}
\todo[inline]{Find citation for children, add priors for $\pi$}
\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
\label{SEIR:table:priors}
\end{table}

\begin{table}[]
\centering
\begin{tabular}{l|lllll}
         & $[0,16)$ & $[16,25)$ & $[25,50)$ & $[50,70)$ & $[70,\infty)$ \\
        \hline
        East Midlands & $\BetaDist(8.5, 301)$ & $\BetaDist(18, 256)$ & $\BetaDist(19, 475)$ & $\BetaDist(20, 493)$ & $\BetaDist(5, 337)$ \\
        East of England & $\BetaDist(7.9, 236)$ & $\BetaDist(21, 248)$ & $\BetaDist(35, 677)$ & $\BetaDist(26, 613)$ & $\BetaDist(5.6, 332)$ \\
        London & $\BetaDist(9.7, 131)$ & $\BetaDist(50, 328)$ & $\BetaDist(154, 1331)$ & $\BetaDist(106, 990)$ & $\BetaDist(7.6, 204)$ \\
        North East & $\BetaDist(6.1, 180)$ & $\BetaDist(12, 145)$ & $\BetaDist(14, 274)$ & $\BetaDist(14, 295)$ & $\BetaDist(4.2, 240)$ \\
        North West & $\BetaDist(12, 252)$ & $\BetaDist(38, 353)$ & $\BetaDist(72, 885)$ & $\BetaDist(44, 695)$ & $\BetaDist(6.3, 264)$ \\
        South East & $\BetaDist(8.9, 404)$ & $\BetaDist(17, 319)$ & $\BetaDist(20, 577)$ & $\BetaDist(19, 586)$ & $\BetaDist(4.4, 356)$ \\
        South West & $\BetaDist(6.9, 415)$ & $\BetaDist(11, 277)$ & $\BetaDist(12, 556)$ & $\BetaDist(11, 559)$ & $\BetaDist(4, 492)$ \\
        West Midlands & $\BetaDist(8, 167)$ & $\BetaDist(19, 151)$ & $\BetaDist(32, 475)$ & $\BetaDist(33, 504)$ & $\BetaDist(5.8, 241)$ \\
        Yorkshire and The Humber & $\BetaDist(9.2, 314)$ & $\BetaDist(19, 242)$ & $\BetaDist(27, 589)$ & $\BetaDist(18, 474)$ & $\BetaDist(4.9, 316)$ \\
    \end{tabular}
\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination. Based on seroprevalence in blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.}
\label{SEIR:table:immunity-prior}
\end{table}
\end{landscape}
